---
title: "Community Biplot"
author: "Ben Weinstein"
date: "July 2, 2015"
output: html_document
---

```{r}
library(vegan)
library(ggbiplot)
library(reshape2)
library(dplyr)
library(tidyr)
source("TraitMappingFunctions.R")
```

Load the site by species and traits data
```{r}
load("sppXsite/mp85bi70.rda")
siteXsppF <- t(sppXsite[,2:(ncol(sppXsite) - 3)])
load("sppXsite/current.rda")
siteXspp <- t(sppXsite[,2:(ncol(sppXsite) - 3)])
traits <- getTraitData()
```
Start with mean traits

For each site (column) get the mean and standard deviation trait value

```{r}

#container for site trait averages
out<-matrix(nrow=ncol(siteXspp),ncol=ncol(traits))
out.sd <- matrix(nrow=ncol(siteXspp), ncol=ncol(traits))
for (x in 1:ncol(siteXspp)){
  #subset site
  site<-siteXspp[,x]
  
  #get the trait matrix for species that are present
  site.trait<-traits[rownames(traits) %in% names(site[site==1]),]
  #mean position
  out[x,]<-apply(site.trait,2,mean)
  out.sd[x,] <- apply(site.trait,2,sd)
}

#name container based on sites
rownames(out)<-colnames(siteXspp)
colnames(out)<-colnames(traits)
rownames(out.sd)<-colnames(siteXspp)
colnames(out.sd)<-colnames(traits)

```

Issue with using the means is that the standard deviation is large (often substantially larger than the mean trait value).

```{r}
head(out.sd)
head(out)
```
Initial biplot

```{r}
#base package
biplot(prcomp(out))

#ggbiplot
ggbiplot((prcomp(out)))
```

Okay so that what it would look like in one time slice. Let's add a new matrix, which is the future siteXspp. 

Repeat trait calculation

```{r}
#container for site trait averages
outF<-matrix(nrow=ncol(siteXsppF),ncol=ncol(traits))

for (x in 1:ncol(siteXsppF)){
  #subset site
  site<-siteXsppF[,x]
  
  #get the trait matrix for species that are present
  site.trait<-traits[rownames(traits) %in% names(site[site==1]),]
  #mean position
  outF[x,]<-apply(site.trait,2,mean)
}

#name container based on sites
rownames(outF)<-colnames(siteXsppF)
colnames(outF)<-colnames(traits)
```

##Approach #1: Bind the dataframes together
```{r}
combin<-rbind(out,outF)
pca<-princomp(combin)
#lets make a grouping variable, so the plot knows which is current and which is future
groups<-c(rep("Current",nrow(out)),rep("Future",nrow(outF)))
ggbiplot(pca,groups=groups)
```

So maybe we want to connect each assemblage by a faint line? Lots of different visualizations we could try

If the two data.frames out (current) and outF (future) are just bound together in a dataframe (combin), and they are ordered in the same manner, then the first row of combin is the current state of the assemblage and the 1+nrow(out) row of combin is the future state of that assemblage. If they are not ordered correctly, you could match off a row name or lookup table by cell number?

### Draw Lines
So as an example, line segements can be drawn in ggplot like this.
```{r}
ggbiplot(pca,groups=groups) + geom_segment(x=0,y=0,xend=0,yend=2,size=1)
```

So the goal is to get the pca coordinates of each pair, and pass them into geom_segement.

Step 1 is to get the correct pair for each line segement. The scores hold the pca loadings that act as coordinates in the figure. We just use the first two PC axis

As an example show the first site
```{r}
#begin coordinates
coords<-pca$scores[c(1,nrow(out)+1),1:2]

#rename to make more logical
colnames(coords)<-c("x","y")
rownames(coords)<-c("Begin","End")

#rearrange the data
mcoords<-melt(coords)
df<-dcast(mcoords,...~Var2+Var1)
df
ggbiplot(pca,groups=groups,scale = 0) + geom_segment(data=df,aes(x=x_Begin,y=y_Begin,xend=x_End,yend=y_End),size=.1,alpha=.5, arrow = arrow(length = unit(0.5,"cm")))
```

Repeat this all current assemblages.

```{r}
#begin coordinates
df<-list()

for (x in 1:nrow(out)){
coords<-pca$scores[c(x,nrow(out)+x),1:2]

#rename to make more logical
colnames(coords)<-c("x","y")
rownames(coords)<-c("Begin","End")

#rearrange the data
mcoords<-melt(coords)
df[[x]]<-dcast(mcoords,...~Var2+Var1)[,-1]
}

df<-rbind.fill(df)
```

plot

```{r}
ggbiplot(pca,groups=groups,scale = 0, varname.size = 5, alpha = 0.05) + geom_segment(data=df,aes(x=x_Begin,y=y_Begin,xend=x_End,yend=y_End),size=.1,alpha=.2, arrow = arrow(length = unit(0.5,"cm"))) + theme_classic() 
```

So It looks as though there is some kind of pattern - mainly that the mean community traits going into the future scenario (MPI-ESM-LR under the 8.5 representative concentration pathway) are those with less extreme values. There is also the cluster of movement towards the bottom left corner (lower mass).